---
title: "CSIT165-Project1"
author: "Jeanine Holscher and Andy Lopez"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include = FALSE}
#### Load necessary packages ####
packages <- c("dplyr", "knitr", "kableExtra", "magrittr", "readr", "geosphere")
install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)

library(dplyr)
library(knitr)
library(kableExtra)
library(magrittr)
library(readr)
library(geosphere)

#### Pull data from files that were obtained with wget
confirmationsURL <- "time_series_covid19_confirmed_global.csv"
confirmations <- read_delim(file=confirmationsURL, delim=",")

deathsURL <- "time_series_covid19_deaths_global.csv"
deaths <- read_delim(file=deathsURL, delim=",")

#### Find the last day listed in Confirmations for future use
latestDate <- names(confirmations)[length(confirmations)]

#### Create new variables from subsets of confirmations and deaths to remove locations with NA or 0 for Latitude for future use 
confClean <- subset(confirmations, !is.na(Lat))
deathsClean <- subset(deaths, !is.na(Lat))

confClean <- subset(confClean, Lat != 0)
deathsClean <- subset(deathsClean, Lat != 0)
```

## Background
The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases. 
*CSIT-165* disseminates data driven analyses to global decision makers.

*CSIT-165* is a conglomerate comprised of two fabricated entities: *Global Health Union (GHU)* and *Private Diagnostic Laboratories (PDL)*. 
Your and your partner's role is to play a data scientist from one of these two entities.

## Data
> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. 
Date fields are stored in MM/DD/YYYY format.

## Project Objectives

### Objective 1
```{r ob1}
#### Create variables for highest confirmation and highest deaths on 1/22/20
confOrigin <- confirmations[which.max(confirmations$"1/22/20"),]
deathOrigin <- deaths[which.max(deaths$"1/22/20"),]

#### Display location with both highest confirmations and deaths, if none qualify display a different message
if ((confOrigin$Lat == deathOrigin$Lat) && (confOrigin$Long == deathOrigin$Long)) {
  print(paste("Predicted origin (Location with highest confirmations and deaths on 1/22/20)"))
  print(paste("Latitude:", confOrigin$Lat, "| Longitude:", confOrigin$Long))
  print(paste("Country/Region:", confOrigin$"Country/Region", "| Province/State:", confOrigin$"Province/State"))
} else {
  print("There is not a location with both the highest confirmations and deaths.")
}
```

### Objective 2
```{r ob2}
#### Initialize counter variable and copy confirmations to use in subsetting
i <- 5
confRecent <- (confirmations)

#### Since we are only counting areas that have wound up with at least one case, subset to remove those who ended at 0 confirmed cases at the last listed date in confirmations
confRecent <- subset(confRecent, ((confRecent)[latestDate]) != 0)

#### Loops through each column and progressively subsets to keep only those rows which are still at 0 confirmations, and breaks the loop when the next subset would result in no data
for (i in 5:length(confRecent))
{
  if(nrow(subset(confRecent, ((confRecent)[i]) == 0)) == 0)
  {
    break
  } else {
    confRecent <- subset(confRecent, ((confRecent)[i]) == 0) 
    recentDate <- names(confRecent)[i+1]
  }

  i <- i + 1
}

#### Displays info for most recent area to have a first confirmed case
print(paste("Most recent area to have a first confirmed case"))
print(paste("Latitude:", confRecent$Lat, "| Longitude:", confRecent$Long))
print(paste("Country/Region:", confRecent$"Country/Region", "| Province/State:", confRecent$"Province/State"))
print(paste("First confirmed case:", recentDate))
```

### Objective 3
```{r ob3}
#### Create variables to hold both the Longitude and Latitude of the origin and recent locations
confOriginLoc <- c(confOrigin$Long, confOrigin$Lat)
confRecentLoc <- c(confRecent$Long, confRecent$Lat)

#### Use distm to find the distance in meters between the two locations and then convert to miles
distanceMeters <- distm(confOriginLoc, confRecentLoc, distGeo)
distanceMiles <- round(distanceMeters * 0.000621371, 2)

#### Display the distance in miles between the two locations
print(paste(confRecent$"Province/State", confRecent$"Country/Region", "is", distanceMiles, "miles away from", confOrigin$"Province/State", confOrigin$"Country/Region"))
```

### Objective 4
```{r ob4}
#### Create new data frame from subset of cleaned confirmations and deaths(without locations with NA or 0 for Latitude)
riskScore <- data.frame("Province/State"=confClean$"Province/State", "Country/Region"=confClean$"Country/Region", "Confirmations"=confClean[latestDate], "Deaths"=deathsClean[latestDate], "Risk Score"=0)

#### Calculate and populate Risk Score field
riskScore[5] <- 100 * (riskScore[4] / riskScore[3])

#### Use colMeans to find the average global risk
avgRisk <- colMeans(riskScore[5], na.rm=TRUE)

#### Subset to find all instances of the lowest and highest risk scores 
lowestRisk <- subset(riskScore, riskScore[5] == (min(riskScore[5], na.rm=TRUE)))
highestRisk <- subset(riskScore, riskScore[5] == (max(riskScore[5], na.rm=TRUE)))

#### Subset again to keep only the lowest risk and highest risk that have the highest confirmations
lowestRisk <- subset(lowestRisk, lowestRisk[3] == (max(lowestRisk[3], na.rm=TRUE)))
highestRisk <- subset(highestRisk, highestRisk[3] == (max(highestRisk[3], na.rm=TRUE)))

#### Display the Global Average risk and info for both lowest and highest risk locations
print(paste("Global Average Risk:", avgRisk))
print(paste("Lowest Risk:",  lowestRisk[1], lowestRisk[2], "| Confirmations:", lowestRisk[3], "| Deaths:", lowestRisk[4], "| Risk:", round(lowestRisk[5], 2)))
print(paste("Highest Risk:", highestRisk[1], highestRisk[2], "| Confirmations:", highestRisk[3], "| Deaths:", highestRisk[4], "| Risk:", round(highestRisk[5], 2)))
```

It might be helpful to calculate risk scores for different areas of the world to be able to assess things like the need for more Covid restrictions / mitigation strategies, travel risks, and need for more testing supplies or health care support. Risk scores assume that people are consistently testing (which also assumes equal access and inclination to testing) and reporting positive tests. They also assume accurate and properly documented reporting of deaths and cause of death. Since this risk score is only taking total confirmations and deaths into account, it is not necessarily representing the current situation accurately - a country could have very few Covid cases currently, but the risk score calculated from the total data could still be very high due to earlier surges.

#### Objective 5
```{r ob5}
#### Make list of unique Countries from cleaned confirmations(without locations with NA or 0 for Latitude)
countryList <- unique(confClean[2])

#### Make data frame structure with country list and blank spots for other data
summedData <- data.frame("Country"=countryList, "Confirmations"=0, "Deaths"=0)

#### Create subsets by country and populate Confirmations
i <- 1

for (i in 1:nrow(summedData))
{
  tempSubset <- subset(confClean, confClean$"Country/Region" == summedData$Country[i])

  summedData$Confirmations[i] <- sum(tempSubset[latestDate], na.rm=TRUE)

  i <- i + 1
}

#### Create subsets by country and populate Deaths
i <- 1

for (i in 1:nrow(summedData))
{
  tempSubset <- subset(deathsClean, deathsClean$"Country/Region" == summedData$Country[i])

  summedData$Deaths[i] <- sum(tempSubset[latestDate], na.rm=TRUE)

  i <- i + 1
}

#### Slice to show only the top five countries for Confirmations and Deaths
topConfirmations <- slice_max(summedData, Confirmations, n = 5)
topDeaths <- slice_max(summedData, Deaths, n = 5)

### Use kable to show both top five tables
knitr::kable(
  topConfirmations,
  "pipe",
  col.names = c("Country", "Confirmations", "Deaths"),
  caption = paste("Countries with Highest Confirmations (as of ", latestDate, ")")
)
knitr::kable(
  topDeaths,
  "pipe",
  col.names = c("Country", "Confirmations", "Deaths"),
  caption = paste("Countries with Highest Deaths (as of ", latestDate, ")")
)
```

### GitHub Link

[Link to Project in GitHub](https://github.com/jholscher1885/CSIT165-Project1)




